---
title: "Introduction to oceancube"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to oceancube}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## One object and five axes

`oceancube` represents rectilinear ocean products with the canonical order
longitude, latitude, depth, time, and variable. The logical contract is always
five-dimensional, including surface fields, which use a singleton depth axis.

```{r create}
library(oceancube)

lon <- c(-80, -79)
lat <- c(-12, -11)
depth <- c(5, 15)
time <- as.Date(c("2020-01-01", "2020-02-01", "2021-01-01", "2021-02-01"))
values <- array(seq_len(2 * 2 * 2 * 4), c(2, 2, 2, 4, 1))
x <- ocean_cube(
  lon, lat, time, values, depth = depth,
  vars = "temperature", units = "degC", source = "offline example"
)
x
summary(x)
```

Memory cubes contain their values. Local NetCDF products can be materialized
with `read_nc()`. The package also has an internal, serializable, read-only
NetCDF descriptor backend; `cube_collect()` turns that representation into an
independent memory cube. No network connection is needed for these workflows.

## Selection and extraction

Slices select stored positions or coordinates, while crops use closed ranges.
Nearest matching is discrete and never interpolates.

```{r selection}
slice <- cube_slice(x, longitude = -79, latitude = -11, by = "value")
crop <- cube_crop(x, longitude = c(-80, -79), depth = c(5, 15))
point <- cube_extract(
  x, longitude = -79, latitude = -11, depth = 5,
  time = time[1], variable = "temperature", by = "value"
)
profile <- cube_extract(
  x, longitude = -79, latitude = -11, time = time[1],
  variable = "temperature", by = "value", mode = "profile"
)
series <- cube_extract(
  x, longitude = -79, latitude = -11, depth = 5,
  variable = "temperature", by = "value", mode = "series"
)
dim(slice$data)
nrow(cube_extract(crop))
point
profile
series
```

`cube_transect()` extracts ordered path-depth combinations. It is experimental
in 0.1.0, so code should name important arguments.

```{r transect}
path <- data.frame(station = c("A", "B"), lon = lon, lat = lat)
cube_transect(
  x, path, lon_col = "lon", lat_col = "lat", id_col = "station",
  depth = 5, time = time[1], variable = "temperature", by = "value"
)
```

## Masks, climatologies, anomalies, and geometry

`cube_mask()` classifies horizontal cell centres against polygons and requires
the optional `sf` package. `cube_aggregate_time()` performs time-only calendar
aggregation with equal observation weights and selective NetCDF reads.
`clim_month()` and `clim_day()` describe recurring temporal baselines; anomalies
retain the cube contract.

```{r temporal}
monthly <- cube_aggregate_time(x, by = "month")
clim <- clim_month(x)
difference <- anom_diff(x, clim)
standardized <- anom_z(x, clim)
clim
dim(monthly$data)
dim(difference$data)
dim(standardized$data)
```

Geometry functions provide cell areas, vertical thicknesses, volumes, and
polygon weights without calculating indicators. See the geometry vignette for
the optional-`sf` workflow.

## Responsibility and limitations

`oceancube` owns representation, local reading, selection, extraction,
transects, masks, climatologies, anomalies, geometry, and weights. Spatial
indicators and inference belong to `spatind`.

Version 0.1.0 is limited to local rectilinear data. It does not write NetCDF,
read OPeNDAP or THREDDS, interpolate, regrid, process curvilinear meshes, or
provide general antimeridian support.
