---
title: "Grid geometry and polygon weights"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Grid geometry and polygon weights}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

`oceancube` exposes geometry primitives for rectilinear grids. Horizontal areas
use geodesic polygon areas; vertical bounds must be explicit or available in
unambiguous cube metadata. These functions do not read ocean values.

```{r cube}
library(oceancube)
x <- ocean_cube(
  lon = c(-80.5, -79.5), lat = c(-12.5, -11.5),
  depth = structure(c(5, 15), units = "m", positive = "down"),
  time = as.Date("2020-01-01"),
  data = array(1:8, c(2, 2, 2, 1, 1)), vars = "temperature"
)
thickness <- cube_layer_thickness(x, depth_bounds = c(0, 10, 20), unit = "m")
thickness
```

Horizontal area, volume, masks, and polygon weights require the optional `sf`
package. The example stays offline and uses a tiny polygon.

```{r sf-geometry}
if (requireNamespace("sf", quietly = TRUE)) {
  area <- cube_cell_area(x, unit = "km2")
  volume <- cube_cell_volume(
    x, depth_bounds = c(0, 10, 20), unit = "km3"
  )
  polygon <- sf::st_as_sf(
    data.frame(id = "region", wkt =
      "POLYGON ((-81 -13, -79 -13, -79 -11, -81 -11, -81 -13))"),
    wkt = "wkt", crs = 4326
  )
  masked <- cube_mask(x, polygon)
  weights_2d <- cube_polygon_weights(x, polygon, id_col = "id")
  weights_3d <- cube_polygon_weights(
    x, polygon, id_col = "id", dimension = "3d",
    depth_bounds = c(0, 10, 20)
  )
  area
  volume
  masked$mask
  weights_2d
  weights_3d
}
```

`cube_mask()` uses cell-centre membership and changes cube values outside the
selected coverage. `cube_polygon_weights()` instead calculates feature-specific
cell intersections and returns self-contained merge keys, areas, fractions,
and—when requested—volumes. It does not aggregate variables or calculate an
indicator. Indicator definitions, normalization, uncertainty, and inference
belong to `spatind`.

The 0.2.0 core supports geographic rectilinear grids and deliberately rejects
unsafe or ambiguous geometry, including unsupported antimeridian cases.
