---
title: "Introduction to oceancube"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to oceancube}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

## One object and five axes

`oceancube` represents rectilinear ocean products with the canonical order
longitude, latitude, depth, time, and variable. The logical contract is always
five-dimensional, including surface fields, which use a singleton depth axis.

```{r create}
library(oceancube)

lon <- c(-80, -79)
lat <- c(-12, -11)
depth <- c(5, 15)
time <- as.Date(c("2020-01-01", "2020-02-01", "2021-01-01", "2021-02-01"))
values <- array(seq_len(2 * 2 * 2 * 4), c(2, 2, 2, 4, 1))
x <- ocean_cube(
  lon, lat, time, values, depth = depth,
  vars = "temperature", units = "degC", source = "offline example"
)
x
summary(x)

validation <- cube_validate(x)
inspection <- cube_inspect(x, missing = "none")
stopifnot(nrow(validation) > 0L)
stopifnot(inherits(inspection, "ocean_cube_inspection"))
```

Memory cubes contain their values. Local NetCDF products can be materialized
with `read_nc()`. The package also has an internal, serializable, read-only
NetCDF descriptor backend; `cube_collect()` turns that representation into an
independent memory cube. No network connection is needed for these workflows.

## Selection and extraction

Slices select stored positions or coordinates, while crops use closed ranges.
Nearest matching is discrete and never interpolates.

```{r selection}
slice <- cube_slice(x, longitude = -79, latitude = -11, by = "value")
crop <- cube_crop(x, longitude = c(-80, -79), depth = c(5, 15))
point <- cube_extract(
  x, longitude = -79, latitude = -11, depth = 5,
  time = time[1], variable = "temperature", by = "value"
)
profile <- cube_extract(
  x, longitude = -79, latitude = -11, time = time[1],
  variable = "temperature", by = "value", mode = "profile"
)
series <- cube_extract(
  x, longitude = -79, latitude = -11, depth = 5,
  variable = "temperature", by = "value", mode = "series"
)
dim(slice$data)
nrow(cube_extract(crop))
point
profile
series
```

`cube_transect()` extracts ordered path-depth combinations. Name important
arguments so matching, tolerance, mode, and distance semantics remain explicit.

```{r transect}
path <- data.frame(station = c("A", "B"), lon = lon, lat = lat)
cube_transect(
  x, path, lon_col = "lon", lat_col = "lat", id_col = "station",
  depth = 5, time = time[1], variable = "temperature", by = "value"
)
```

## Static visualization

The five visualization functions return `ggplot` objects and preserve the
scientific cube. They select stored cells without interpolation, smoothing, or
imputation.

```{r visualization}
plots <- list(
  map = viz.map(x, "temperature", time = time[1], depth = 5),
  section = viz.section(
    x, "temperature", time = time[1], latitude = lat[1]
  ),
  profile = viz.profile(
    x, "temperature", longitude = lon[1], latitude = lat[1], time = time[1]
  ),
  transect = viz.transect(
    x, path, "temperature", time = time[1], depth = 5,
    lon_col = "lon", lat_col = "lat", id_col = "station"
  ),
  timeseries = viz.timeseries(
    x, "temperature", longitude = lon[1], latitude = lat[1], depth = 5
  )
)
stopifnot(all(vapply(plots, inherits, logical(1), what = "ggplot")))
```

## Masks, climatologies, anomalies, and geometry

`cube_mask()` classifies horizontal cell centres against polygons and requires
the optional `sf` package. `cube_aggregate_time()` performs time-only calendar
aggregation with equal observation weights and selective NetCDF reads.
`cube_climatology()` describes recurrent day, month, or meteorological-season
baselines through equally weighted period-year replicates and bounded NetCDF
reads. Its `data` contains the climatological mean and `climatology$sd` contains
aligned sample SD. `clim_month()` and `clim_day()` remain compatibility
wrappers. New workflows use the canonical climatology directly with
`cube_anomaly()`; legacy anomaly wrappers delegate to the same engine.
`cube_trend()` adds descriptive per-cell OLS slopes against actual elapsed
historical time. It accepts raw, aggregated, and anomaly cubes, but rejects
recurrent climatology pseudo-time.

```{r temporal}
monthly <- cube_aggregate_time(x, by = "month")
clim_cube <- suppressWarnings(cube_climatology(x, by = "month"))
clim <- suppressWarnings(clim_month(x))
difference <- cube_anomaly(x, clim_cube, type = "difference")
standardized <- cube_anomaly(x, clim_cube, type = "z")
legacy_difference <- anom_diff(x, clim)
clim_cube
clim
dim(monthly$data)
dim(difference$data)
dim(standardized$data)
trend <- cube_trend(x, time_unit = "year")
dim(trend$data)
```

Geometry functions provide cell areas, vertical thicknesses, volumes, and
polygon weights without calculating indicators. See the geometry vignette for
the optional-`sf` workflow.

## Responsibility and limitations

`oceancube` owns representation, local reading, selection, extraction,
transects, masks, climatologies, anomalies, descriptive per-cell temporal
slopes, geometry, and weights. Spatial indicators, trend inference, change
points, and regimes belong to downstream analysis such as `spatind`.

The current core is limited to local rectilinear data. It does not write
NetCDF, read OPeNDAP or THREDDS, interpolate, regrid, process curvilinear
meshes, perform Sen/Mann--Kendall inference, detect breakpoints or regimes, or
provide general antimeridian support.
